\section{Suite Inventory}

The full benchmark family has six suites:
\begin{itemize}
  \item Suite A: consequence-to-action.
  \item Suite B: reafferent source attribution.
  \item Suite C: re-engagement under world change.
  \item Suite D: long-horizon moved bottleneck.
  \item Suite E: tool commitment and repair.
  \item Suite F: structure-compatible OOD.
\end{itemize}

The main paper centers Suite C and Suite D/E because they are the strongest
current anchors. Suites A/B/F should remain in appendix or a suite-family
overview unless expanded into their own benchmark cards.

\section{Suite C Gate Definitions}

Source:
\path{experiments/world_responds/results/suite_c_reengagement_2026_07_06.md}
and
\path{experiments/world_responds/results/suite_c_neural_transfer_2026_07_06.md}.

\begin{table*}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
Gate & Meaning & Current status \\
\midrule
C1 & baseline silence replication & pass \\
C2 & selective first-shift re-engagement & pass \\
C3 & affected-component recovery & pass \\
C4 & no false calm & pass \\
C5 & cost-aware inquiry & pass \\
C6 & second-shift reopenability & pass \\
N1 & stale/wrong/suppressed learned controls fail & pass in learned transfer \\
\bottomrule
\end{tabular}
\caption{Suite C gates.}
\end{table*}

\section{Suite D/E Gate Surface}

Source: \path{experiments/long_horizon_bottleneck/BENCHMARK_CARD.md}.

The benchmark should report:
\begin{itemize}
  \item closed-loop final accuracy,
  \item moved-slot specificity and rank,
  \item visible-control null behavior,
  \item hidden localization over model/layer/token-position grids,
  \item fixed-action counterfactual localization,
  \item causal patch value-prefix effects,
  \item prompt-family robustness,
  \item API black-box prompt-family and external-stress behavior.
\end{itemize}

\section{Release Schema}

The machine-readable summary must contain:
\begin{verbatim}
benchmark, suite, run, minimum_pass_rule,
score_axes, gates, baselines, artifacts,
allowed_claim, non_claims
\end{verbatim}

Rows should be JSONL. Summary JSON should be public-safe and should not include
secrets, uncurated provider transcripts, or local-only Modal payloads.

\section{Known Negatives and Boundaries}

Known negatives are valuable and should not be hidden:
\begin{itemize}
  \item signal-layer cooling can create false calm in Suite C,
  \item matched-random inquiry at the same budget fails selectivity/recovery,
  \item stale and wrong stress signals fail the learned Suite C transfer,
  \item API dispatch failure for one tested OpenAI condition is sparse and
  localized, not broad,
  \item Suite F remains a companion OOD package rather than a complete benchmark
  release inside this paper.
\end{itemize}
