\documentclass[10pt,twocolumn]{article}
\usepackage[margin=0.75in]{geometry}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\title{Teacher-Free Adaptive Inquiry Under World Change}
\author{Anonymous Authors}
\date{}
\begin{document}
\maketitle

\begin{abstract}
Suite C previously showed that a hand-specified policy and then a
teacher-trained probe head can re-engage after world change. This paper reports
the next finite step: a linear inquiry policy selected by teacher-free
cross-entropy-method reward search. The training objective uses downstream
recovery, selective re-engagement, second-shift reopenability, probe cost, and
anti-cheat controls; it does not consume hand-policy labels, teacher actions, or
teacher probabilities. On held-out seeds the policy passes C1-C6 plus T1/N1,
and a 64-seed bootstrap replication preserves the effect.
\end{abstract}

\section{Introduction}
The reviewer objection is direct: imitation of a hand-designed inquiry law is
weaker than teacher-free adaptive inquiry. The result here removes direct
teacher supervision while keeping the same minimum pass rule: final recovery is
not sufficient unless inquiry is selective, fresh, cost-aware, reopenable, and
robust to anti-cheat controls.

\section{Suite C Task}
Describe two world shifts, affected and unaffected buckets, probe cost, recovery,
false calm, and second-shift reopenability.

\section{Teacher-Free Training}
The implemented policy is a linear probe head over the existing observable
Suite C feature surface: perceived error, perceived surprise, jumps relative to
a lagging baseline, effort, improvement, time since probe, recent probe rate,
source identity, and bucket index. Cross-entropy-method search selects
parameters using only downstream reward terms for recovery, selectivity,
reopenability, probe cost, no-false-calm behavior, and control failure.

\section{Gates and Controls}
The accepted policy must pass C1-C6 plus T1 and N1. T1 audits that no teacher
labels, teacher actions, or teacher probabilities enter the training loss. N1
requires stale-signal, wrong-signal, signal-suppression, and matched-random
controls to fail in the intended ways.

\section{Results}
The first held-out run passes all gates: final affected MAE 0.095, recovery rate
1.000, first-shift selectivity 12.500, second-shift reopenability 11.312, and
22.0 probes. Matched random at the same budget reaches selectivity 0.833;
stale-signal recovery is 0.000; wrong-signal selectivity is 0.000; and
suppressed-signal final MAE is 0.464.

A wider 64-seed bootstrap replication preserves the conclusion. Final affected
MAE is 0.095 with 95\% CI [0.095, 0.096]; recovery rate is 1.000 [1.000,
1.000]; selectivity lift over matched random is 11.292 [10.976, 11.596]; and
suite pass rate is 1.000 [1.000, 1.000].

\section{Limitations}
Finite simulator, no biological or consciousness claim, no production-agent
claim, long-horizon/open-agent transfer remains open. The source-identity
feature is still privileged; a paper-ready follow-up should replace it with a
learned or noisy source estimator.

\end{document}
